The Unicode Cookbook \linebreak for Linguists

### 2018-06-06 book/chapters/orthography_profiles.tex
deals with paperhivers through chapter 2

Given these goals, Unicode Locales (see
Chapter~\ref{the-unicode-approach}) might seem like the ideal orthography
profiles. However, there are various practical obstacles preventing the use of
such locale descriptions in the daily linguistic practice, namely: (i) the
XML structure is too verbose to easily and quickly produce or correct manually,
(ii) locale descriptions are designed for a wide scope of information (like date
formats or names of weekdays) most of which is not applicable for documenting
writing systems, and (iii) most crucially, even if someone made the effort to
produce a technically correct locale description for a specific source at hand,
then it is well-nigh impossible to deploy the description. This is because a locale
description has to be submitted to and accepted by the Unicode Common Locale
Data Repository. The repository is (rightly so) not interested in descriptions
that only apply to a limited set of sources (e.g.~descriptions for only a single
dictionary).

>>>>

Given these goals, Unicode Locales (see
Chapter~\ref{the-unicode-approach}) might seem like the ideal orthography
profiles. However, there are various practical obstacles preventing the use of
Unicode Locales in the daily linguistic practice, namely: (i) the
XML structure is too verbose to easily and quickly produce or correct manually,
(ii) Unicode Locales are designed for a wide scope of information (like date
formats or names of weekdays) most of which is not applicable for documenting
writing systems, and (iii) most crucially, even if someone made the effort to
produce a technically correct Unicode Locale for a specific source at hand,
then it is well-nigh impossible to deploy the description. This is because a locale
description has to be submitted to and accepted by the Unicode Common Locale
Data Repository. The repository is (rightly so) not interested in descriptions
that only apply to a limited set of sources (e.g.~descriptions for only a single
dictionary).

### 2018-06-06 book/chapters/unicode_approach.tex
deals with paperhivers through chapter 2

The Unicode Standard represents a massive step forward because it aims to
eradicate the distinction between universal (ASCII) versus language-particular
(font) by adding as much language-specific information as possible into the
universal standard. However, there are still language/resource-specific
specifications necessary for the proper usage of Unicode, as will be discussed
below. Within the Unicode structure many of these specifications can be captured
by so-called \textsc{locale descriptions}, so we are moving to a new distinction
of universal (Unicode Standard) versus language-particular (locale description).
The major gain is much larger compatibility on the universal level (because
Unicode standardizes a much greater portion of writing system diversity), and
much better possibilities for automated processing on the language-particular
level (because locale descriptions are machine-readable specifications).

@@

These properties contain the basic information of the Unicode Standard and they
are necessary to define the correct behavior and conformance required for
interoperability in and across different software implementations (as defined in
the Unicode Standard Annexes). The character properties assigned to each code
point are based on each character's behavior in real-world writing
traditions. For example, the corresponding lowercase character to \uni{0047} is
\uni{0067}.\footnote{Note that the relation between uppercase and lowercase is in
many situations much more complex than this, and Unicode has further
specifications for those cases.} Another use of properties is to define the
script of a character.\footnote{The Unicode Glossary defines the term \textsc{script} as
a ``collection of letters and other written signs used to represent textual
information in one or more writing systems. For example, Russian is written with
a subset of the Cyrillic script; Ukrainian is written with a different subset.
The Japanese writing system uses several scripts.''} In practice, script is
simply defined for each character as the explicit \textsc{script} property in
the Unicode Character Database.

@@

The (default) \textsc{grapheme clusters} are locale-independent graphemes,
i.e.~they always apply when a particular combination of characters occurs
independent of the writing system in which they are used. These character
combinations are defined in the Unicode Standard as functioning as one
\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
``A minimum unit of text in relation to a particular text process, in the
context of a given writing system. In general, the mapping between text elements
and code points is many-to-many.''} The simplest example of a grapheme cluster
is a base character followed by a letter modifier character. For example, the
sequence <n> + <\dia{0303}> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
user-perceived character in writing systems like that of Spanish. In effect, what the
user perceives as a single character actually involves a multi-code-point
sequence. Note that this specific sequence can also be represented with a single
so-called \textsc{precomposed code point}, the \textsc{latin small letter n with
tilde} at \uni{00F1}, but this is not the case for all multi-code-point
character sequences. A solution to the problem of multiple encodings for the
same text element was developed early on in the Unicode Standard. It is called 
\textsc{canonical equivalence}, e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} 
should in all situations be treated identically to the precomposed \uni{00F1}. By doing so, 
Unicode can also support special or precomposed characters in legacy character sets. 
To determine canonical equivalence, the Unicode Standard offers different kinds of normalization to either 
decompose precomposed characters (called \textsc{NFD} for \textsc{normalization form
canonical decomposition}) or to combine sequences of code points into precomposed 
characters (called \textsc{NFC} for \textsc{normalization form canonical composition}).\footnote{See the 
Unicode Standard Annex \#15, Unicode Normalization Forms (\url{http://unicode.org/reports/tr15/}), 
which provides a detailed description of normalization algorithms and illustrated examples.} 
In current software development practice, NFC seems to be preferred in most situations
and is widely proposed as the preferred canonical form. We discuss Unicode normalization 
in detail in Section \ref{pitfall-canonical-equivalence}.

More difficult for text processing, because less standardized, is what the
Unicode Standard terms \textsc{tailored grapheme clusters}.\footnote{\url{http://unicode.org/reports/tr29/}} 
Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do
not function as text elements in all situations. Examples include the sequence
<c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme
clusters are \textsc{tailored} in the sense that they must be specified on a
language-by-language or writing-system-by-writing-system basis. They are also 
grapheme clusters in these orthographies for processes such as collation (i.e.\ sorting).\footnote{\url{https://www.unicode.org/glossary/#collation}}

The Unicode Standard provides technical specifications for creating locale specific data
in so-called \textsc{Unicode Locales}, i.e.~specifications 
that define a set of language-specific elements (e.g.~tailored grapheme
clusters, collation order, capitalization-equivalence), as well as other special
information, like how to format numbers, dates, or currencies. Locale
descriptions are saved in the \textsc{Common Locale Data Repository
(CLDR)},\footnote{More information about the CLDR can be found here:
\url{http://cldr.unicode.org/}.} a repository of
language-specific definitions of writing system properties, each of which
describes specific usages of characters. Each locale can be encoded in a
document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
format and vocabulary for the exchange of structured locale data. Unicode Locale
Descriptions allow users to define language- or even resource-specific writing
systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
system} only very loosely, as it is not a central concept in the Unicode
Standard. A writing system is, ``A set of rules for using one or more scripts to
write a particular language. Examples include the American English writing
system, the British English writing system, the French writing system, and the
Japanese writing system.''} However, there are various drawbacks of locale
descriptions for the daily practice of linguistic work in a multilingual setting.

>>>>

The Unicode Standard represents a massive step forward because it aims to
eradicate the distinction between universal (ASCII) versus language-particular
(font) by adding as much language-specific information as possible into the
universal standard. However, there are still language/resource-specific
specifications necessary for the proper usage of Unicode, as will be discussed
below. Within the Unicode structure many of these specifications can be captured
by so-called \textsc{Unicode Locales}, so we are moving to a new distinction
of universal (Unicode Standard) versus language-particular (Unicode Locale).
The major gain is much larger compatibility on the universal level (because
Unicode standardizes a much greater portion of writing system diversity), and
much better possibilities for automated processing on the language-particular
level (because Unicode Locales are machine-readable specifications).

@@

These properties contain the basic information of the Unicode Standard and they
are necessary to define the correct behavior and conformance required for
interoperability in and across different software implementations (as defined in
the Unicode Standard Annexes). The character properties assigned to each code
point are based on each character's behavior in real-world writing
traditions. For example, the corresponding lowercase character to \uni{0047} is
\uni{0067}.\footnote{Note that the relation between uppercase and lowercase is in
many situations much more complex than this, and Unicode has further
specifications for those cases.} Another use of properties is to define the
script of a character.\footnote{The Glossary of Unicode Terms defines the term \textsc{script} as
a ``collection of letters and other written signs used to represent textual
information in one or more writing systems. For example, Russian is written with
a subset of the Cyrillic script; Ukrainian is written with a different subset.
The Japanese writing system uses several scripts.''} In practice, script is
simply defined for each character as the explicit \textsc{script} property in
the Unicode Character Database.

@@

The (default) \textsc{grapheme clusters} are locale-independent graphemes,
i.e.~they always apply when a particular combination of characters occurs
independent of the writing system in which they are used. These character
combinations are defined in the Unicode Standard as functioning as one
\textsc{text element}.\footnote{The Glossary of Unicode Terms defines \textsc{text element} as:
``A minimum unit of text in relation to a particular text process, in the
context of a given writing system. In general, the mapping between text elements
and code points is many-to-many.''} The simplest example of a grapheme cluster
is a base character followed by a letter modifier character. For example, the
sequence <n> + <\dia{0303}> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
user-perceived character in writing systems like that of Spanish. In effect, what the
user perceives as a single character actually involves a multi-code-point
sequence. Note that this specific sequence can also be represented with a single
so-called \textsc{precomposed code point}, the \textsc{latin small letter n with
tilde} at \uni{00F1}, but this is not the case for all multi-code-point
character sequences. A solution to the problem of multiple encodings for the
same text element was developed early on in the Unicode Standard. It is called 
\textsc{canonical equivalence}, e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} 
should in all situations be treated identically to the precomposed \uni{00F1}. By doing so, 
Unicode can also support special or precomposed characters in legacy character sets. 
To determine canonical equivalence, the Unicode Standard offers different kinds of normalization to either 
decompose precomposed characters (called \textsc{NFD} for \textsc{normalization form
canonical decomposition}) or to combine sequences of code points into precomposed 
characters (called \textsc{NFC} for \textsc{normalization form canonical composition}).\footnote{See the 
Unicode Standard Annex \#15, Unicode Normalization Forms (\url{http://unicode.org/reports/tr15/}), 
which provides a detailed description of normalization algorithms and illustrated examples.} 
In current software development practice, NFC seems to be preferred in most situations
and is widely proposed as the preferred canonical form. We discuss Unicode normalization 
in detail in Section \ref{pitfall-canonical-equivalence}.

More difficult for text processing, because less standardized, is what the
Unicode Standard terms \textsc{tailored grapheme clusters}.\footnote{\url{http://unicode.org/reports/tr29/}} 
Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do
not function as text elements in all situations. Examples include the sequence
<c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme
clusters are \textsc{tailored} in the sense that they must be specified on a
language-by-language or writing-system-by-writing-system basis. They are also 
grapheme clusters in these orthographies for processes such as collation (i.e.\ sorting).\footnote{\url{https://www.unicode.org/glossary/\#collation}}

The Unicode Standard provides technical specifications for creating locale specific data
in so-called \textsc{Unicode Locales}, i.e.~specifications 
that define a set of language-specific elements (e.g.~tailored grapheme
clusters, collation order, capitalization-equivalence), as well as other special
information, like how to format numbers, dates, or currencies. Locale
descriptions are saved in the \textsc{Common Locale Data Repository
(CLDR)},\footnote{More information about the CLDR can be found here:
\url{http://cldr.unicode.org/}.} a repository of
language-specific definitions of writing system properties, each of which
describes specific usages of characters. Each locale can be encoded in a
document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
format and vocabulary for the exchange of structured locale data. Unicode Locale
Descriptions allow users to define language- or even resource-specific writing
systems or orthographies.\footnote{The Glossary of Unicode Terms defines \textsc{writing
system} only very loosely, as it is not a central concept in the Unicode
Standard. A writing system is, ``A set of rules for using one or more scripts to
write a particular language. Examples include the American English writing
system, the British English writing system, the French writing system, and the
Japanese writing system.''} However, Unicode Locales have various drawbacks 
for the daily practice of scientific linguistic research in a multilingual setting.
